
## Math 365 / Comp 365: Activity 14
## Polynomial Interpolation

```{r,message=FALSE}
require(Matrix)
```

**Question 1**

a. Write a function `Vandermonde(x)` that takes a sequence of interpolation points (just the $x_i$ values) and returns the Vandermonde matrix for that sequence. Hint: what does the following code do and why?
```{r}
outer(1:3,1:3)
outer(1:3,1:3,`^`)
```

```{r}
Vandermonde = function(x) {
  n = length(x)
  V = outer(x, 0:(n-1), `^`)
  return(V)
}
```


b. Try running `V=Vandermonde(1:5)`. What is the condition number of $V$?

```{r}
V = Vandermonde(1:5)
print(V)
# Condition number of a matrix
Cond = function(A,p=2) {
  if (p == 2) {  # by default use the 2-norm
    s = svd(A)$d
    s = s[s>0]
    return(max(s)/min(s))
  }
  if (p == 1) {  # use the 1 norm
    Ainv = solve(A)
    return(max(colSums(abs(A)))*max(colSums(abs(Ainv))))
  }
  if (p == 'I') {   # use the infinity norm
    Ainv = solve(A)
    return(max(rowSums(abs(A)))*max(rowSums(abs(Ainv))))
  }
}

Cond(V)
```

The condition number of V is 26169.69, which is not good.

**Question 2**

Write a second function `VandermondeInterpolator(x,y)` that takes a sequence of interpolation points and values (both the $x_i$ and $y_i$ values) and i) returns a list containing both the coefficients of the polynomial interpolant (ordered from the constant term up to the degree n-1 term) and a function for evaluating the polynomial interpolant for the given values, and ii) plots both the points and interpolation function on the same graph. You can use the built-in `solve` function to solve the $Ax=b$ problem.

```{r}
Horner = function(coeffs,x,b=rep(0,length(coeffs))) {
  y = coeffs[length(coeffs)]
  for (i in (length(coeffs)-1):1) {
    y = y*(x-b[i])+coeffs[i] 
  }
  return(y)
}


VandermondeInterpolator = function(x,y) {
  V = Vandermonde(x)
  C = solve(V, y)
  H = Horner(C, x)
  plot(x, y)
  
  return(List(C=C, H=H))
}
```


**Question 3**

Consider the points (-2,0), (-1,14), (2,-4), (3,10). Find the degree 3 (i.e., the highest order term is $cx^3$) interpolating polynomial p(x) for these points in two different ways. 

a) Use your `VandermondeInterpolator` function from Question 2. 

b) By hand, find the interpolating polynomial using Newton's divided differences.  You need to multiply it out to see that you are getting the same answer as above.

Note: You can also find the same interpolating polynomial using Lagrange interpolation. I won't ask you to do this, but you should at least look at Lagrange interpolation and see how the algebra would yield the same solution. 
