---
title: "Exam 1"
author: Lawson Busch
output: 
  html_document:
    toc: true
    toc_float: true
---




\
\


## Getting started

```{r}
library(dplyr)
library(infer)
library(broom)

# 85 complete sample counties
not_the_data <- read.csv("https://www.macalester.edu/~ajohns24/data/counties_253.csv")
counties     <- na.omit(not_the_data)
```




\
\


## EXERCISE 1


My loop to calculate LOOCV for my model:
```{r}
loocv = rep(0, nrow(counties))
for(i in 1:(nrow(counties))) {
  train <- counties[-i,]
  test <- counties[i,]
  
  model_train <- lm(growth ~ ., train)
  model_test <- predict(model_train, newdata = data.frame(counties[i,]))
  
  test_observed <- test$growth
  test_residuals <- test_observed - model_test
  
  mspe_test = mean(test_residuals^2)
  loocv[i] = mspe_test
}
mean(loocv)
```

And checking using the cv.glm function:
```{r}
library(boot)
counties_model_glm <- glm(growth ~ ., counties, family = "gaussian")
counties_model_cv <- cv.glm(counties, counties_model_glm)
counties_model_cv$delta
```

We see that my calculated loocv is correct.
\
\


## EXERCISE 2


My calculated loocv error is 837.0871, which matches the loocv error calculated by the cv.glm function.


\
\


## EXERCISE 3




```{r}
counties <- counties %>%
  mutate(population = density*area) %>%
  mutate(spending_per_person = fed_spending / population)
```







\
\


## EXERCISE 4




```{r}
median(counties$spending_per_person)
```


Bootstrapping

First take 1000 samples:
```{r}
set.seed(253)
counties_resamples <- rep_sample_n(counties, size = nrow(counties), replace = TRUE, reps = 1000)
```

This gives us 1000 different "samples" of our data. So to find the new sample median for spending_per_person, we can first get the median for each of our resamples. 

```{r}
counties_medians <- counties_resamples %>%
  group_by(replicate) %>%
  summarize(mdn = median(spending_per_person))
```

We now have 1000 sample medians. We can then calculate and interpret a 95% confidence interval for the true population median by using the quantile function:

```{r}
quantile(counties_medians$mdn, c(0.025, 0.975))
```

Or we can say that we are 95% confident that the true spending_per_person population median (as calculated by our 1000 bootstrap samples) falls between 7.784047 and 8.977165 dollars per person.

\
\




## EXERCISE 5


We do not have enough evidence to support the claim that the government provides counties with more than 8 dollars per resident because our confidence interval falls between 7.78 and 8.97. So while the majority of our confidence interval's range is above 8 dollars, we cannot say that the government provides counties with more than 8 dollars per resident, because our confidence interval covers an area less than 8 dollars per resident.








\
\


**THE END**

Submit your knit HTML to Moodle!




