---
title: "Using Logistic Regression for Classification"
author: "Lawson Busch"
date: "10/16/2018"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r}
library(ggplot2)
library(gridExtra)
```


To load the data:
```{r}
vertebra <- read.csv("https://www.macalester.edu/~ajohns24/data/vertebral_column.csv")

# Check out the data
head(vertebra, 3)
##   pelvic_incidence pelvic_tilt lumbar_angle sacral_slope pelvic_radius
## 1            63.03       22.55        39.61        40.48         98.67
## 2            39.06       10.06        25.02        29.00        114.41
## 3            68.83       22.22        50.09        46.61        105.99
##   spondylolisthesis_grade class
## 1                   -0.25     1
## 2                    4.56     1
## 3                   -3.53     1

# Summarize the vertebral class
table(vertebra$class)
## 
##   0   1 
## 100 210
```

## Warm-Up

To make the logistic regression model of class by pelvic_incidence:
```{r}
class_mod <- glm(class ~ pelvic_incidence, vertebra, family="binomial")
summary(class_mod)
```

So the model for our vertebra class in log(odds) form is:

$$log(odds) = -2.36+0.05*pelvicIncidence$$

To interpret this coefficient, we can say that for every extra 1 point increase in p_i, we expect log(odds of irreg) to increase by 0.053774. In a more meaningful interpretation we can convert this coefficient to odds for to get the multiplicative change in odds, where $odds = exp(0.053774) = 1.055246$. Or that for every time we increase p_i by one, we expect the odds to increase by a factor of 1.055246.

And the model for our vertebra class in probability form is:

$$probability(irregeular) = \frac{e^{-2.36+0.05*pelvicIncidence}}{e^{-2.36+0.05*pelvicIncidence} + 1}$$

```{r}
#Plot of the raw data
g1 <- ggplot(vertebra, aes(x = pelvic_incidence, fill = factor(class))) +
    geom_density(alpha = 0.5)

# Plot of the model
g2 <- ggplot(vertebra, aes(x = pelvic_incidence, y = class)) +
    geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) + 
    labs(y = "P(irregular)")

grid.arrange(g1, g2, ncol = 2)
```

So to predict P(irregular) if pelvic_incidence = 100:

```{r}
(exp(-2.362786+0.053774*100))/(exp(-2.362786+0.053774*100) +1)
```

Or we can say patients with a pelvic_incidence of 100 have a 95.3% chance of being classified as irregular. This can be verrified using the predict funtion.

```{r}
predict(class_mod, newdata = data.frame(pelvic_incidence = 100), type = "response")
```

