---
title: "Team Project 2"
author: Raven McKnight, Lawson Busch, Ximena Silvia
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r message = FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(maps)
library(usmap)
library(ggrepel)
library(geosphere)
library(leaps)
```
```{r message = FALSE}
flights  <- read_csv("https://www.macalester.edu/~ajohns24/data/flights.csv")
airlines <- read_csv("https://www.macalester.edu/~dshuman1/data/112/flights/airlines.csv")
airports <- read_csv("https://www.macalester.edu/~dshuman1/data/112/flights/airports.csv")
```


## Part 1

```{r}
# typical difference between actual & scheduled arrive times
mean(na.omit(flights$ARRIVAL_DELAY))
min(na.omit(flights$ARRIVAL_DELAY))
max(na.omit(flights$ARRIVAL_DELAY))
```

We see that the mean arrival delay for the flights in our data set is 4.36 minutes, the minimum arrival delay is arriving 80 minutes early, and the maximum arrival delay is arrival delay is 1463 minutes late.

We want to see which flights are on time, delayed, and cancelled within our data set. We will define "ontime" as within 15 minutes of the predicted arrival time. 

```{r}
# new categorical variable
flights <- flights %>%
  mutate(delay_cancel = if_else(is.na(ARRIVAL_DELAY), "cancelled", (if_else(ARRIVAL_DELAY<=abs(15), "ontime", "delayed"))))
```

The table below displays the number of flights which are cancelled, delayed, and on time as well as the proportion of the flights in each group. 

```{r}
prop <- flights %>%
  count(delay_cancel)
prop <- prop%>%
  mutate(prop = n/nrow(flights))

prop
```

We can see all these flights on the following plot:

```{r}
# Create dataset with ony Origin and Destination airport codes
flights_mod <- flights %>%
  arrange(flights$ORIGIN_AIRPORT) %>%
  select(ORIGIN_AIRPORT, DESTINATION_AIRPORT)

# create new dataset without the airports with numerical airport codes because they do not have corresponding values in airports$IATA_CODE
flights_mod <- flights_mod[8404:100000,]
# remoe the duplicate flights
flight_mod <- unique(flights_mod[, 1:2])
# modify dataset to be able to store the longitude and latitude values
flight_mod <- flight_mod %>%
  mutate(Lat_from = NA, Lon_from = NA, Lat_to = NA, Lon_to = NA)

#fix all the airports that were missing coordinates
airports$LATITUDE[314] = 29.9544
airports$LONGITUDE[314] = -81.3429

airports$LATITUDE[97] = 30.3578
airports$LONGITUDE[97] = -85.7938

airports$LATITUDE[235] = 26.8234
airports$LONGITUDE[235] = -80.1387

# parse through the data and assign the appropriate latitude and longitude values based on the IATA codes and information in airports dataset for both Origin Airport and Destination Airport
for (i in 1:dim(flight_mod)[1]){
  for (j in 1:322){
    if (flight_mod$ORIGIN_AIRPORT[i] == airports$IATA_CODE[j]){
      flight_mod$Lat_from[i] = airports$LATITUDE[j]
      flight_mod$Lon_from[i] = airports$LONGITUDE[j]
    }
  }
  for (k in 1:322){
    if (flight_mod$DESTINATION_AIRPORT[i] == airports$IATA_CODE[k]){
      flight_mod$Lat_to[i] = airports$LATITUDE[k]
      flight_mod$Lon_to[i] = airports$LONGITUDE[k]
    }
  }
  
}

# Create an environment to map all the data points onto
# xlim and ylim values set to include Hawaii and Alaska to the west and Puerto Rico to the east 
usmap1 <- map("world", fill=T, col="grey8", bg="grey15",
              xlim=c(-170.0,-65.0) ,ylim=c(0,72.0)
              )
# Map all the airports onto the usmap1 environment 
points(airports$LONGITUDE,airports$LATITUDE, pch=3, cex=0.1, col="chocolate1")
# title for our plot
title(main = "All unique flight patterns in data set", col.main = "white")
# parsed through all the data points and create curved splines connecting Origin and Departure airports
for (x in 1: 4259){
  inter <- gcIntermediate(c( flight_mod$Lon_from[x],flight_mod$Lat_from[x]), 
                          c( flight_mod$Lon_to[x], flight_mod$Lat_to[x]), 
                          n = 150)
  lines(inter, lwd=0.06, col="turquoise2")
}
```

## Part 2

We want to implement a model that predicts whether or not a flight will be on time, cancelled or delayed. To do this, we will first take a smaller sample size of our original data set.

```{r}
set.seed(253)
small_flights <- sample_n(flights, size = 1000, replace = FALSE)
```

Our first insight was to drop any predictors from the original flight data set that were redundant or didn't factor into our investigation. This included variables such as ARRIVAL_DELAY, which is just ARRIVAL_TIME - SCHEDULED_ARRIVAL, and the plane's tail number. This shrunk down our number of predictors significantly. We also decided to remove airports, as it counted for over 300 seperate categorical predictors, and slowed down our computation time significantly. 

Next, we decided that the best tool would predict whether a flight would be delayed or cancelled based solely on information that is available pre-flight, such as day, time, airline, etc. This meant dropping all of the predictors from the data set that had to do with delayed time. We decided to drop these predictors because 1) they are only available post flight, so you can't use them to predict if a flight will be delayed before take off, and 2) if there is delay data, obviously the flight was delayed. 

This left us with a set of predictors that only included what we referred to as "ticket" information, which is all the information present on an airplane ticket. However, we quickly realized that this information could not be used to predict delays, as none of the predictors were strong enough to categorize a flight as on time, delayed or cancelled. To demonstrate this, we have included plots of our flights data sample. Below is a plot of the delay_cancel variable by day of week, as a percentage of the data:

```{r}
library(ggplot2)
ggplot(small_flights, aes(x=DAY_OF_WEEK, fill=delay_cancel)) + geom_bar(position="fill")
```

What we can see from the above plot is that regardless of the day of the week, roughly the same percentage of flights are on time, cancelled, or delayed. This demonstrates that the day of the weak is a weak predictor for whether or not a flight will be on time. We can see a similar pattern with airline's below:

```{r}
ggplot(small_flights, aes(x=AIRLINE, fill=delay_cancel)) + geom_bar(position="fill")
```

We can also see that time of day is not a particularly strong predictor either. While it is somewhat stronger that airline or day of the week, we can still see that it is not a great predictor for a flight being on time.

```{r}
ggplot(small_flights, aes(x=SCHEDULED_TIME, fill=delay_cancel)) + geom_histogram(color="white")
```

We quickly realized that all of our "ticket information" predictors followed this pattern, making the a model using solely these predictors practically useless. To counter this, we decided to include one more predictor: TAXI_OUT. We decided on TAXI_OUT because it is a preflight variable that does not necessarily correspond to a delay (as every flight needs a taxi), and it enabled us to make a tree that also incorporated the other predictors. 

```{r}
flightsM <- small_flights %>% 
  select(c(-YEAR, -DAY, -FLIGHT_NUMBER, -TAIL_NUMBER, -ORIGIN_AIRPORT, -DESTINATION_AIRPORT, -DEPARTURE_TIME, -DEPARTURE_DELAY, -WHEELS_OFF, -ELAPSED_TIME, -AIR_TIME, -WHEELS_ON, -TAXI_IN, -ARRIVAL_TIME, -ARRIVAL_DELAY, -DIVERTED, -CANCELLATION_REASON, -CANCELLED, -AIR_SYSTEM_DELAY, -SECURITY_DELAY, -AIRLINE_DELAY, -LATE_AIRCRAFT_DELAY, -WEATHER_DELAY)) 
```

To make our tree:

```{r}
library(rpart)
library(randomForest)

tree_1 <- rpart(delay_cancel ~., flightsM)
tree_1
plot(tree_1, margin = 0.2)
text(tree_1, cex = 0.8)
```

From the diagram we can see that the TAXI_OUT variable dominates this tree. If your taxi takes 37.5 or more minutes, your flight is likely to be delayed. However, it also allows us to see cutoffs for other variables, such as flights under 1,390 miles are likely to be ontime, and that flights in the early months of the year (the winter) are more likely to be delayed than flights in later months. Overall, we can see that the best flights short flights later in the year, during weekdays, that depart in the morning with short taxi times, which logically makes sense. Further, it is important to note that our model never predicts a flight as cancelled. This is likely in part due to only 1.6% of flights being cancelled, and none of our predictors having a strong relationship with a flight being cancelled (as discussed above).

Lastly, we want to evaluate the accuracy of this model. To do this we will take another random testing sample (with a different seed) of our original flights data. We will then see how well our model performed.

```{r}
set.seed(2000)
small_flights_test <- sample_n(flights, size = 1000, replace = FALSE)

pred_1 <- predict(tree_1, newdata = small_flights_test, type = "class")    
true <- small_flights_test$delay_cancel

res <- table(pred_1, true)
res
```

As discussed above, we can see that our model does not predict cancelled flights at all. It's overall accuracy is:

```{r}
sum(diag(res)) / nrow(small_flights_test)
```

Or that it accurately predicts 81.3% of flights. Breaking this down further:

```{r}
delayed_acc <- 25 / (25 + 134)
ontime_acc <- 788 / (33 + 788)
delayed_acc
ontime_acc
```

We can see that our model predicts 0% of cancelled flights correctly, 15.7% of delayed flights correctly and 95.9% of ontime flights correctly. In other words, our model is extremely good at telling you if a flight will be on time, and terrible at predicting if it will be delayed or cancelled.

Below, we can see the flight patterns of each unique delayed, cancelled, and on time flight:

```{r}
# Create new data set 
flights_mod_1 <- flights %>%
  arrange(ORIGIN_AIRPORT) %>%
  select(delay_cancel, ORIGIN_AIRPORT, DESTINATION_AIRPORT)
# Include all the data points except the data with numeric airport codes
flights_mod_1 <- flights_mod_1[8404:100000,]
# Remove duplicates and then NA values
flight_mod_1<- unique(flights_mod_1[, 1:3])
flight_mod_1 <- na.omit(flight_mod_1)
# Modify dataset to be able to store coordinate values
flight_mod_1 <- flight_mod_1 %>%
  mutate(Lat_from = NA, Lon_from = NA, Lat_to = NA, Lon_to = NA)

# Include coordinate values from airports dataset 
for (i in 1:8625){
  for (j in 1:322){
    if (flight_mod_1$ORIGIN_AIRPORT[i] == airports$IATA_CODE[j]){
      flight_mod_1$Lat_from[i] = airports$LATITUDE[j]
      flight_mod_1$Lon_from[i] = airports$LONGITUDE[j]
    }
  }
  for (k in 1:322){
    if (flight_mod_1$DESTINATION_AIRPORT[i] == airports$IATA_CODE[k]){
      flight_mod_1$Lat_to[i] = airports$LATITUDE[k]
      flight_mod_1$Lon_to[i] = airports$LONGITUDE[k]
    }
  }
  
}

# Seperate data based on the delay_cancel value
late_flight <- flight_mod_1 %>%
  filter(delay_cancel == "delayed")
ontime_flight <- flight_mod_1 %>%
  filter(delay_cancel == "ontime")
cancel_flight <- flight_mod_1 %>%
  filter(delay_cancel == "cancelled")

# Create environment like before
usmap1 <- map("world", fill=T, col="grey8", bg="grey15",
              xlim=c(-170.0,-65.0) ,ylim=c(0,72.0)
              )
# Airports
points(airports$LONGITUDE,airports$LATITUDE, pch=3, cex=0.1, col="chocolate1")
# Information for viewer/reader
title(main = "Flight patterns of on time, delayed, and cancelled flights", col.main = "white")
legend("topright", legend = c("delayed", "ontime", "cancelled"), fill = c("red", "blue", "green"), border = "black", bg = "grey8", text.col = "white")
# Plot all the paths, changing color depending on delay_cancel value
for (x in 1:3371){
  inter <- gcIntermediate(c(late_flight$Lon_from[x],late_flight$Lat_from[x]), 
                          c( late_flight$Lon_to[x], late_flight$Lat_to[x]), 
                          n = 150)
  lines(inter, lwd=0.2, col= "red")
}
for (y in 1: 4165){
  inter1 <- gcIntermediate(c( ontime_flight$Lon_from[y],ontime_flight$Lat_from[y]), 
                          c( ontime_flight$Lon_to[y], ontime_flight$Lat_to[y]), 
                          n = 150)
  lines(inter1, lwd=0.1, col= "blue")
}
for (z in 1: 1065){
  inter <- gcIntermediate(c( cancel_flight$Lon_from[z],cancel_flight$Lat_from[z]), 
                          c( cancel_flight$Lon_to[z], cancel_flight$Lat_to[z]), 
                          n = 150)
  lines(inter, lwd=0.06, col= "green")
}
```


\
\
\
\
\
\



## Part 3

Not only would we like to predict which flights will be delayed, cancelled, or on time but we would like to know how long of a delay each flight is likely to have. To do this, we will swap out the delay_cancel variable for the ARRIVAL_DELAY, and form a model to predict ARRIVAL_DELAY.

```{r}
flightsN <- small_flights %>% 
  select(c(-YEAR, -DAY, -FLIGHT_NUMBER, -TAIL_NUMBER, -ORIGIN_AIRPORT, -DESTINATION_AIRPORT, -DEPARTURE_TIME, -DEPARTURE_DELAY, -WHEELS_OFF, -ELAPSED_TIME, -AIR_TIME, -WHEELS_ON, -TAXI_IN, -ARRIVAL_TIME, -DIVERTED, -CANCELLATION_REASON, -CANCELLED, -AIR_SYSTEM_DELAY, -SECURITY_DELAY, -AIRLINE_DELAY, -LATE_AIRCRAFT_DELAY, -WEATHER_DELAY, -delay_cancel))
```

To determine which predictors to use in our model, we will do best subsets selection. This will create all possible models with our set of predictors, and turn out the best possible model for each corresponding number of predictors. 

```{r}
best_sub <- regsubsets(ARRIVAL_DELAY ~., flightsN, method = "forward", really.big = TRUE, nvmax = 8)

# Store the summary information
best_summary <- summary(best_sub)
best_summary
```

Next, we can plot the adjusted R squared value for each returned model, and use this to select our number of predictors.

```{r}
plot_data <- data.frame(size = 1:8, adjr2 = best_summary$adjr2)
ggplot(plot_data, aes(x = size, y = adjr2)) + 
    geom_point() + 
    geom_line() + 
    labs(x="subset size", y="Adjusted R-squared")
```

From the above plot, we can see that the adjusted R squared value begins to round out at 6 predictors, at an adjusted R squared of 0.08. This implies that our model should use the six predictor model from the best subsets.



